---
title: "Practice Day 1 Answers"
---

This is the notebook fellows will work through on the first day of the Center for Health Journalism Hands-On R course.

You will be using daily weather summaries that have been downloaded  from [Climate Data Online](https://www.ncei.noaa.gov/cdo-web/datasets). You can choose to work with data from Arkansas, California or Texas. The explanations use Texas, but you may use one of the other files available.

## Setup

This is the section for libraries.

```{r}
library(tidyverse)
library(janitor)
```

## Import

Follow the directions in the book to import your data. 

```{r}
tx_raw <- read_csv("data/tx.csv") |> clean_names()

tx_raw |> head()
```

And now glimpse the data

```{r}
tx_raw |> glimpse()
```

## Select

Use this section to create a "tight" version of your data with only the columns you need.

```{r}
tx_tight <- tx_raw |> 
  select(
    name,
    date,
    prcp,
    snow,
    snwd,
    tmax,
    tmin
  )

tx_tight |> glimpse()
```

## Arrange

Find the coldest day, warmest day, most snow, most rain.

## Mutate

Create a year value.

## Filter

Choose a city, save as new. Find days with snow.

## Group and Summarize: Count

Find the years with the most snow days. Find the years with the most 100+ degree days.

## Group and Summarize: Math

Years with most rain. Years with most snow.

## Recode

Go back and shorten your station names.

```{r}
tx_tight |> 
  group_by(name) |> 
  slice_max(tmax)

tx_tight |> 
  group_by(name) |> 
  slice_min(tmin)
```



