---
title: "Introduction"
---

The idea is this ...

There would be pre-written code, with explanations of what it does. Fellows would work through this notebook, running pre-written code and adding new code of their own in mini "on your own" quests.

Practice is in a different [notebook](p1-intro.qmd).

## Libraries

Talk about them.

```{r}
#| label: setup
#| message: false
#| echo: fenced

library(tidyverse)
library(janitor)
```

## Importing

A lot of talk about how this works ...

```{r}
tx <- read_csv("data-raw/tx.csv") |> clean_names()

tx |> glimpse()
```

No OYO for importing since we don't want to do it twice.


### OYO: Import

Insert a new code block and import the weather data from a different state.


> The challenge here is the page will stop rendering if they get an error, but hopefully we'll work through that since it will be an issue in any event.


## Select columns

Info about doing it. Example with Texas.

### OYO: Select

Directions to do on your own with another state.




